---
title: "Combined stations and predators"
format:
  html:
    code-fold: true
editor: visual
output-dir: "docs"
---

```{r}
#| label: setup
#| include: false

library(lubridate)
library(sf)
library(targets)
library(tidyverse)

tar_load(c("effort_sf", "predators_agg", "predators_stations", 
           "station_effort", "stations_ice", "zoop", "zoop_sf"),
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

## Sightings and stations

Visual description of line transect and point station aggregation, for supplemental.

```{r}
#| label: fig-aggregation
#| fig-cap: "Biophysical observations (e.g., macrozooplankton net tows and CTD casts) collected at stations (yellow point) were associated with visual surveys for predators collected along line transects. All visual surveys conducted within 15 km (shaded circle) and 3 days were aggregated with station observations to form sites. In this example, the station was sampled on August 15, 2016 and visual surveys were condected within 15 km of the station twice: on August 15 (red) and on August 23 (blue). Only the visual surveys collected on August 15 were associated with this site."
#| warning: false

points_to_line <- function(pts) {
  pts %>% 
    group_by(segmentid) %>% 
    summarize(begin = paste0(first(processID), first(interval)),
              end = paste0(last(processID), last(interval)),
              lbl = format(first(UTC_start), "%b %d %Y"),
              do_union = FALSE) %>% 
    mutate(segmentid = factor(segmentid, labels = lbl)) %>% 
    st_cast("LINESTRING")
}

tow_origin <- as.POSIXct("1899-12-31", tz = "UTC")
station_prj <- slice(zoop_sf, 25) %>% 
  st_transform(ant_proj()) %>% 
  mutate(tow.UTC = date.tow + (start.time.UTC - tow_origin),
         lbl = format(tow.UTC, "%b %d"))
station_buffer <- st_buffer(station_prj, 15e3)
station_bbox <- st_bbox(station_buffer) %>% 
  expand_bbox(1.5)
transect_prj <- effort_sf %>% 
  filter(between(UTC_start, 
                 station_prj$tow.UTC - days(3),
                 station_prj$tow.UTC + days(3))) %>% 
  st_transform(ant_proj()) %>% 
  st_intersection(station_buffer) %>% 
  mutate(lag = UTC_start - lag(UTC_start, default = UTC_start[1]),
         segmentid = factor(cumsum(lag > hours(6))))
transect_line <- transect_prj %>% 
  points_to_line()
transect_prj2 <- effort_sf %>% 
  filter(year == station_prj$Year) %>% 
  st_transform(ant_proj()) %>% 
  st_crop(station_bbox) %>% 
  mutate(lag = UTC_start - lag(UTC_start, default = UTC_start[1]),
         segmentid = factor(cumsum(lag > hours(6))))
transect_line2 <- transect_prj2 %>% 
  points_to_line()

transect_endpoints <- transect_prj2 %>% 
  filter(paste0(processID, interval) %in% c(transect_line2$begin, 
                                            transect_line2$end))

ant_basemap() +
  geom_sf(data = station_buffer,
          fill = "grey50",
          alpha = 0.1,
          linetype = "dashed") +
  geom_sf(aes(color = segmentid),
          transect_line2,
          linetype = "dotted",
          linewidth = 1) +
  geom_sf(aes(color = segmentid),
          transect_line,
          linewidth = 1) +
  geom_sf(data = station_prj,
          shape = 23,
          size = 4,
          fill = "gold") +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  coord_ant(station_bbox) +
  theme(legend.position = c(0.03, 0.98),
        legend.justification = c(0, 1),
        legend.title = element_blank())

```

Station locations on predator (five most abundant) count maps. Spatially restricted to station transects.

```{r}
#| label: combined_map
#| warning: false
#| column: screen
#| out-width: 100%
#| fig-width: 6
#| fig-dpi: 600

top_five <- predators_agg %>% 
  group_by(species) %>% 
  summarize(total_count = sum(count)) %>% 
  arrange(desc(total_count)) %>%
  slice(1:5)

topfive_sf <- predators_agg %>% 
  semi_join(top_five, by = "species") %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))

topfive_stars <- rasterize_counts(topfive_sf, res = 2.5e4)

zoop_sf <- latlon_to_sf(zoop, 
                        coords = c("dec.longitude", "dec.latitude")) %>% 
  st_transform(ant_proj()) %>% 
  rename(year = Year)
zoop_buffers_20k <- st_buffer(zoop_sf, 20e3)

map_lim <- st_bbox(zoop_sf) %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)

ant_basemap() +
  stars::geom_stars(aes(fill = count), data = topfive_stars) +
  geom_sf(aes(color = time.of.day), zoop_sf, size = 0.2) +
  scale_x_continuous(breaks = c(-65, -60, -55)) +
  scale_y_continuous(breaks = c(-66, -64, -62, -60)) +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                      trans = scales::pseudo_log_trans(),
                      breaks = c(0, 25, 100, 400, 1000, 4000),
                      na.value = NA) +
  scale_color_manual(values = tod_pal()) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  coord_ant(map_lim) +
  facet_grid(species ~ year) +
  expand_limits(fill = 0) +
  theme(axis.text.x = element_text(size = 7),
        legend.position = "bottom",
        legend.title = element_blank())
```

As above, but for species of interest (Gentoo Penguins, Blue Petrels, Antarctic Terns, Leopard Seals, and Killer Whales).

```{r}
#| label: combined_map_soi
#| warning: false
#| column: screen
#| out-width: 100%
#| fig-width: 6
#| fig-dpi: 600

soi <- c("GEPN", "BLPT", "ANTE", "LESE", "KIWH")

soi_sf <- predators_agg %>% 
  filter(species %in% soi) %>% 
  latlon_to_sf(coords = c("lon_mean", "lat_mean"))

soi_stars <- rasterize_counts(soi_sf, res = 2.5e4)

ant_basemap() +
  stars::geom_stars(aes(fill = count), data = soi_stars) +
  geom_sf(aes(color = time.of.day), zoop_sf, size = 0.2) +
  scale_x_continuous(breaks = c(-65, -60, -55)) +
  scale_y_continuous(breaks = c(-66, -64, -62, -60)) +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                      trans = scales::pseudo_log_trans(),
                      breaks = c(0, 25, 100, 400, 1000, 4000),
                      na.value = NA) +
  scale_color_manual(values = tod_pal()) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  coord_ant(map_lim) +
  facet_grid(species ~ year) +
  expand_limits(fill = 0) +
  theme(axis.text.x = element_text(size = 7),
        legend.position = "bottom",
        legend.title = element_blank())
```

## Aggregated ice observations

```{r}
ice_qa <- zoop_sf %>% 
  right_join(stations_ice, by = "amlr.station") %>%
  distinct(amlr.station, ice_type, ice_intervals, ice_type_freq) %>%
  mutate(n_obs = ice_intervals,
         n_mode = round(ice_type_freq, digits = 0))
```

Each station was associated with the modal ice type (multi-year, first-year, thin, or open) from the underway observations. This figure shows how frequent the modal ice type was at each station. The two grey lines represent 1:1 and 2:1. Stations falling on the upper line had *uniform* ice category observations i.e., every interval within 15 km had the same ice category. Stations falling below the lower line had *highly mixed* ice category observations i.e., the most frequent ice category was observed at less than half the intervals within 15 km of the station.

The frequency of the modal ice type was 90% or greater for 57.8% of stations. The modal ice type was \<50% frequency at just 10 stations (\<5% of total stations).

```{r}
#| label: underway-ice

ggplot(ice_qa, aes(n_obs, n_mode)) +
  geom_abline(slope = c(1, 0.5), intercept = 0, color = "grey50") +
  geom_point(alpha = 0.5) +
  labs(x = "Count of intervals with ice category observation",
       y = "Count of intervals with most frequent ice category") +
  coord_fixed() +
  facet_wrap(~ ice_type, ncol = 2) +
  theme_classic()
```

```{r}
#| label: ice-map
#| column: screen
#| out-width: 100% 

ice_sf <- station_effort %>% 
  select(amlr.station, year) %>% 
  right_join(stations_ice, by = "amlr.station")

ant_basemap() +
  geom_sf(aes(color = ice_type), 
          ice_sf,
          size = 0.5) +
  scale_color_brewer("Ice category", palette = "YlGnBu") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  facet_wrap(~ year) +
  theme(legend.position = "left")

```

```{r}
ice_type_grand_total_row <- stations_ice %>% 
  mutate(YEAR = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  count(ice_type) %>% 
  pivot_wider(names_from = "ice_type", 
              values_from = "n", 
              values_fill = 0) %>% 
  mutate(TOTAL = OPEN + THIN + FY + MY,
         across(-TOTAL, \(x) x / TOTAL),
         YEAR = "Total")

stations_ice %>% 
  mutate(YEAR = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  count(YEAR, ice_type) %>% 
  pivot_wider(names_from = "ice_type", 
              values_from = "n", 
              values_fill = 0) %>% 
  mutate(TOTAL = OPEN + THIN + FY + MY,
         across(c(OPEN, THIN, FY, MY), \(x) x / TOTAL)) %>% 
  rbind(ice_type_grand_total_row) %>% 
  mutate(across(-c(TOTAL, YEAR), scales::percent)) %>% 
  knitr::kable()

stations_ice %>% 
  mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  count(year, ice_type) %>% 
  group_by(year) %>% 
  mutate(frac = n / sum(n)) %>% 
  ungroup() %>% 
  mutate(ice_type = fct_rev(ice_type)) %>% 
  ggplot(aes(year, frac, fill = ice_type)) +
  geom_col() +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous("Sea ice composition", labels = scales::percent) +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank())
stations_ice %>% 
  mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  group_by(year, ice_type) %>% 
  summarize(mean_cov = mean(ice_coverage) / 10,
            n = n(),
            .groups = "drop_last") %>% 
  mutate(n = sum(n),
         se_cov = sqrt(mean_cov * (1 - mean_cov) / n)) %>% 
  ungroup() %>% 
  mutate(ice_type = fct_rev(ice_type)) %>% 
  ggplot(aes(year, mean_cov, fill = ice_type)) +
  geom_errorbar(aes(ymin = mean_cov - se_cov, ymax = mean_cov + se_cov),
                width = 0.2,
                color = "grey50",
                position = "dodge") +
  geom_point(size = 2, 
             shape = 21, 
             color = "grey50", 
             position = position_dodge(width = 0.2)) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(limits = c(NA, 1),
                     labels = scales::percent) +
  theme_minimal() +
  theme()

stations_ice %>% 
  mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  group_by(year) %>% 
  summarize(mean = mean(ice_coverage),
            sd = sd(ice_coverage),
            .groups = "drop") %>% 
  mutate(mean_sd = str_glue("{format(mean, digits = 2)} \U00B1 {format(sd, digits = 2)}")) %>% 
  knitr::kable()
```

## Effort

### Number of stations

Total of `r n_distinct(predators_stations$amlr.station)` stations. Breakdown by year:

```{r}
predators_stations %>% 
  distinct(amlr.station) %>% 
  mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  count(year)
```

### Line transects

```{r}
survey_km <- predators_stations %>% 
  distinct(amlr.station, survey_km) %>% 
  pull(survey_km)
mean_km <- format(mean(survey_km), digits = 3) 
sd_km <- format(sd(survey_km), digits = 3) 
```

`r mean_km` $\pm$ `r sd_km` km of survey effort associated with each station.

### Ice and effort combined

```{r}
percent <- partial(scales::percent, accuracy = 0.1)

ice_type_lookup = c(
  OPEN = "Open water",
  THIN = "Thin ice",
  FY = "First-year ice",
  MY = "Multi-year ice"
)

ice_table <- stations_ice %>% 
  mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  group_by(year, ice_type) %>% 
  summarize(n_year_type = n(),
            ice_mean = mean(ice_coverage / 10),
            .groups = "drop_last") %>% 
  mutate(n_year = sum(n_year_type),
         ice_se = sqrt(ice_mean * (1 - ice_mean) / n_year)) %>% 
  ungroup() %>% 
  transmute(year,
            ice_type = ice_type_lookup[ice_type],
            ice_val = str_glue("{n_year_type} ({percent(ice_mean)} \U00B1 {percent(ice_se)})")) %>% 
  pivot_wider(names_from = ice_type, 
              values_from = ice_val, 
              values_fill = "0") %>% 
  rbind(
    stations_ice %>% 
      mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
      group_by(ice_type) %>% 
      summarize(n_type = n(),
                ice_mean = mean(ice_coverage / 10)) %>% 
      mutate(ice_se = sqrt(ice_mean * (1 - ice_mean) / sum(n_type)),
             ice_val = str_glue("{n_type} ({percent(ice_mean)} \U00B1 {percent(ice_se)})")) %>% 
      transmute(year = "Total", ice_type = ice_type_lookup[ice_type], ice_val = str_glue("{n_type} ({percent(ice_mean)} \U00B1 {percent(ice_se)})")) %>% 
      pivot_wider(names_from = ice_type, 
              values_from = ice_val)
  )

effort_table <- predators_stations %>%
  distinct(amlr.station, survey_km) %>% 
  mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
  group_by(year) %>% 
  summarize(`Stations sampled` = n(),
            mean_km = mean(survey_km),
            sd_km = sd(survey_km),
            `Effort per station (km)` = str_glue("{sprintf('%0.1f', mean_km)} \U00B1 {sprintf('%0.1f', sd_km)}")) %>% 
  select(year, `Stations sampled`, `Effort per station (km)`) %>% 
  rbind(
    predators_stations %>% 
      distinct(amlr.station, survey_km) %>% 
      mutate(year = str_extract(amlr.station, "AMLR([0-9]{4}).*", 1)) %>% 
      summarize(
        year = "Total",
        `Stations sampled` = n(),
        `Effort per station (km)` = str_glue("{sprintf('%0.1f', mean(survey_km))} \U00B1 {sprintf('%0.1f', sd(survey_km))}")
      )
  )

effort_table %>% 
  left_join(ice_table, by = "year") %>% 
  knitr::kable() %>% 
  kableExtra::add_header_above(c(" ", "Survey effort" = 2, "Ice conditions" = 4))
```
