---
title: "Predator clusters"
format:
  html:
    code-fold: true
    toc: true
editor: visual
output-dir: "docs"
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

library(cluster)
library(ggmosaic)
library(ggtree)
library(patchwork)
library(sf)
library(targets)
library(tidyverse)
library(vegan)

tar_load(c("predators_stations", "predators_abundant", 
           "sightings_clust", "sightings_mtx", "sightings_dist", 
           "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

## Predator clusters

Following the methods of Dietrich et al. (2021), clustering analysis is restricted to species sighted near 5% or more of stations. This avoids granting undue leverage to rare species. `r n_distinct(predators_abundant$species)` species retained (i.e., species above red line).

```{r}
#| label: abundant_predators

total_stations <- n_distinct(zoop_sf$amlr.station)
pred_frac <- predators_stations %>%
  as_tibble() %>% 
  group_by(amlr.station, species) %>% 
  summarize(count_station = sum(count),
            .groups = "drop") %>% 
  group_by(species) %>%
  summarize(total_count = sum(count_station),
            stations_present = n(),
            station_frac = stations_present / total_stations) %>%
  arrange(desc(station_frac))
cutoff_row <- sum(pred_frac$station_frac >= 0.05)
  
frac_fn <- scales::label_percent(accuracy = 0.1)
pred_frac %>% 
  set_names(c("Species", "Ind", "Stations", "Fraction")) %>% 
  mutate(Fraction = frac_fn(Fraction)) %>% 
  knitr::kable(align = "lrr") %>% 
  kableExtra::kable_styling("striped",
                            full_width = FALSE) %>% 
  kableExtra::row_spec(cutoff_row, 
                       extra_css = "border-bottom: 2px dashed red")

```

## Optimal number of clusters

We use the gap statistic [@tibshirani2001] to choose the optimal number of clusters ($k$). The gap statistic ($f$) is a goodness of clustering measure. The authors recommended choosing the smallest $k$ such that $f(k) \geq f(k+1) - s(k+1)$ where $s$ is the standard error of $f$. According to this rule, the predators are best described by two clusters (indicated in blue). However, $f(2)$ is only *slightly* greater than the threshold $f(3) - s(3)$, and $f$ is determined stochastically, so depending on the iteration the rule may choose $k=3$. Based on our prior knowledge of three predator guilds (pagophilic, open water, and intermediate), we'll present the results for $k=3$ here and re-run the analysis for $k=2$ in the supplemental.

```{r}
#| label: optimal-clusters

cluster_fn <- function(x, k) {
  bray_dist <- function(x) {
    vegdist(x, method = "bray")
  }
  list(cluster = cutree(hclust(bray_dist(x), method = "ward.D2"), k = k))
}
set.seed(139)
sightings_gap <- suppressWarnings(
  clusGap(sightings_mtx, 
                   FUN = cluster_fn,
                   K.max = 10, 
                   B = 500,
                   verbose = FALSE)
)
sightings_gap_df <- as_tibble(sightings_gap$Tab) %>% 
  mutate(
    k = seq_along(logW),
    is_optimum = k == maxSE(gap, SE.sim, method = "Tibs2001SEmax")
  )
ktibs <- filter(sightings_gap_df, is_optimum) %>% pull(k)
tibs_thr <- with(sightings_gap_df, gap[ktibs + 1] - SE.sim[ktibs + 1])
ggplot(sightings_gap_df, aes(x = k, y = gap)) +
  geom_line() +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim,
                      color = is_optimum)) +
  geom_segment(x = ktibs, xend = ktibs + 1, 
               y = tibs_thr, yend = tibs_thr,
               color = "blue",
               linetype = "dashed") +
  scale_color_manual(values = c(`TRUE` = "blue",
                                `FALSE` = "black")) +
  theme_classic() +
  theme(legend.position = "none")
```

Hierarchical clustering of stations by predator community. Note: `r nrow(zoop_sf)` stations were included in the zooplankton clustering, but only `r length(sightings_clust$labels)` stations were used for predator clustering.

```{r}
#| label: predator_dendro

# Cut predators into five clusters
clust <- cutree(sightings_clust, 3)
g <- split(names(clust), clust)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

p %>% 
  groupClade(pred_clust_mrca, group_name = "Predator cluster") + 
  aes(color = `Predator cluster`) +
  layout_dendrogram() +
  theme_dendrogram()
```

How do the predator clusters compare to the zooplankton clusters?

```{r}
#| label: predator_mosaic
#| warning: false

# Join with zooplankton clusters
both_clust <- zoop_sf %>% 
  transmute(amlr.station, 
            zoop_clust = factor(`Winter Cluster factor`)) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

# Mosaic plot
cluster_pal = c(`1` = "#96B687",
                `2a` = "#9CB2C8",
                `2b` = "#CAB3CC",
                `3a` = "#D29A97",
                `3b` = "#E8AD5C")
ggplot(both_clust) +
  geom_mosaic(aes(x = product(zoop_clust, pred_clust), 
                  fill = zoop_clust)) +
  geom_mosaic_text(aes(x = product(zoop_clust, pred_clust),
                       label = after_stat(.wt)),
                   as.label = TRUE) +
  scale_fill_manual(values = cluster_pal) +
  labs(x = "Predator cluster",
       y = "Zooplankton cluster",
       fill = "Zooplankton cluster") +
  theme_classic()
```

The above mosaic plot shows the distribution of stations by predator and zooplankton clusters. For example, the bottom left rectangle indicates 34 stations were assigned to zooplankton cluster 1 and predator cluster 1. Rectangles color-coded by zooplankton cluster using the color palette from Dietrich et al. 2021 Fig 3a.

Upon visual inspection of the mosaic plot, there appears to be some general correlation. Predator cluster 1 associates somewhat indiscriminately with zooplankton clusters 2a-3b. Predator cluster 2 associates strongly with zooplankton cluster 2a, as well as with 2b to a lesser extent. Predator cluster 3 associates with zooplankton clusters 3a and 3b. Zooplankton cluster 1 seems to be the "background". 

For context, zooplankton cluster 1 was mostly salps, clusters 2a and 2b were small krill and amphipods (mostly offshore), and clusters 3a and 3b were large krill and *Metridia* (mostly in the Bransfield Strait).

## Indicator species

Which species best represent each cluster? Using Dufrene-Legendre indicator species analysis. Apply indicator value cutoff at 25%.

```{r}
#| label: indicator_species

species_wide <- predators_abundant %>% 
  pivot_wider(names_from = species,
              values_from = count_norm,
              values_fill = 0) %>% 
  column_to_rownames("amlr.station")
cluster_indval <- labdsv::indval(species_wide, clust)
indval_mtx <- cluster_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  result <- ifelse(x >= 0.25, 
         str_glue("{rownames(indval_mtx)} ({round(x * 100)}%)"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()

```

### Cluster descriptions

* Cluster 1 = open ocean? (ANFU)
* Cluster 2 = marginal ice zone (SNPT, ANPT, FUSE)
* Cluster 3 = pagophilic (ADPN, CRSE, LESE)

## Cluster maps

What geospatial patterns do we see in the predator communities? The marginal ice zone community was mostly observed just north of the South Shetland islands, occasionally farther north beyond Elephant Island. The open ocean community occupied the northeastern portion of the study area in 2012, 2014, and 2015. In 2013 it contracted westward and in 2016 it was entirely absent. The pagophilic predator community dominated the Bransfield Strait itself, expanding north towards Elephant Island in 2013 and 2016.

```{r}
#| label: cluster_maps

pred_clust_sf <- tibble(
  amlr.station = names(clust),
  pred_clust = factor(clust,
                      labels = c("Open", "Marginal", "Pagophilic"))
) %>% 
  left_join(select(zoop_sf,
                   amlr.station,
                   geometry,
                   Year),
            by = "amlr.station") %>% 
  st_as_sf() %>% 
  st_transform(ant_proj())

map_lim <- st_bbox(pred_clust_sf) %>% 
  project_bbox() %>% 
  expand_bbox(factor = 1.2)

ant_basemap(map_lim) +
  geom_sf(aes(color = pred_clust), pred_clust_sf, size = 2, shape = 15) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = c(-60, -55)) +
  scale_y_continuous(breaks = c(-63, -62, -61, -60)) +
  scale_color_brewer(palette = "Dark2") +
  coord_ant(map_lim) +
  theme(legend.position = "bottom")
```

## NMDS

How many NMDS axes should we use? Scree plot

```{r}
#| label: nmds-scree

nmds_stress <- tibble(
  k = 1:6,
  stress = map_dbl(k, function(k) {
    metaMDS(
      sightings_mtx, 
      distance = "bray",
      k = k,
      trymax = 100,
      trace = FALSE
    )$stress
  })
)

ggplot(nmds_stress, aes(k, stress)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = c(0.2, 0.1, 0.05), linetype = "dotted") +
  scale_x_continuous(breaks = nmds_stress$k) +
  expand_limits(y = 0) +
  theme_classic()

```

3 axes for NMDS looks acceptable. How does ordination distance compare to dissimilarity (Shepard diagram)?

```{r}
#| label: nmds-shepard

sightings_nmds <- metaMDS(
  sightings_mtx, 
  distance = "bray",
  k = 3,
  trymax = 100,
  trace = FALSE
)

nmds_isoreg <- isoreg(sightings_nmds$diss, sightings_nmds$dist)

nmds_shepard <- tibble(
  diss = sightings_nmds$diss,
  dist = sightings_nmds$dist,
  mono = nmds_isoreg$yf
)

nmds_spear <- cor(nmds_shepard$diss,
                  nmds_shepard$dist,
                  method = "spearman")
nmds_label <- str_glue(
  "Non-metric fit, R^2 = {round(1 - sightings_nmds$stress^2, 3)}\n",
  "Linear fit, R^2 = {round(nmds_spear^2, 3)}\n",
  "k = {sightings_nmds$ndim}\n",
  "Stress = {round(sightings_nmds$stress, 3)}"
)

ggplot(nmds_shepard, aes(diss, dist)) +
  geom_point(size = 0.75, alpha = 0.1) +
  geom_line(aes(y = mono), color = "red") +
  annotate("text", x = 0.05, y = 2.9, label = nmds_label,
           hjust = 0, vjust = 1) +
  expand_limits(x = 0, y = 0) +
  labs(x = "Observed dissimilarity",
       y = "Ordination distance") +
  theme_classic()

```

NMDS plots, color-coded by predator cluster.

```{r}
#| label: loading

nmds_df <- sightings_nmds$points %>% 
  as_tibble(rownames = "amlr.station") %>% 
  rename_with(~ paste0("N", .x), starts_with("MDS")) %>% 
  right_join(pred_clust_sf %>% 
               as_tibble() %>% 
               select(amlr.station, pred_clust),
             by = "amlr.station")

ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

ggplot(nmds_df, aes(NMDS1, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

ggplot(nmds_df, aes(NMDS2, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()
```

## Environmental loadings

I compared the drivers of community structure two ways: physics (hydrology and ice coverage) and zooplankton (community assignments). Here's what makes jumps out at me:

* Predator cluster associations with ice kind of track with what we expected. The pagophilic predator cluster correlates with first-year ice, open ocean predators with open ice, marginal ice predators with thin ice. Weirdly, multi-year ice falls among the open ocean predators, but only two stations were classified as multi-year ice so it's probably a small sample size thing.
* Predator cluster 1 (open water) and 2 (marginal ice zone) are associated with higher temperatures and zooplankton cluster 2 (small euphausiids, amphipods, etc). They differ in the open water predator cluster was found in areas with greater phaeopigment and the marginal ice cluster in areas with greater chlorophyll-a.
* Predator cluster 3 (pagophilic) is associated with higher salinity, deeper mixed layer, and macrozooplankton cluster 3 (big euphausiids, Metridia, etc)
* Macrozooplankton cluster 1 (salps etc) evenly distributed across predator clusters

```{r}
#| label: env-loadings

station_ice <- predators_stations %>% 
  distinct(amlr.station, ice_type, ice_coverage) %>% 
  as_tibble()

# Evironmental variables
station_env <- zoop_sf %>% 
  select(amlr.station, zuml_m, avg.temp, avg.salinity, Integ.chla.100m, 
         Integ.phae.100m, TOD_2levels_civil, Year, 
         zoop = `Winter Cluster factor`) %>% 
  as_tibble() %>% 
  left_join(station_ice, by = "amlr.station") %>% 
  mutate(zoop = fct_collapse(zoop,
                             `2` = c("2a", "2b"),
                             `3` = c("3a", "3b")),
         Year = factor(Year)) %>% 
  slice(map_int(rownames(sightings_mtx),
                ~ which(.x == amlr.station))) %>% 
  select(-amlr.station, -geometry)

# Fit multiple regression model
pred_envfit <- envfit(sightings_nmds, 
                      station_env, 
                      permutations = 999, 
                      choices = 1:3,
                      na.rm = TRUE)

# Extract envfit vectors and factors
envfit_vecs <- scores(pred_envfit, 
                      display = "vectors",
                      choices = 1:3) %>% 
  as_tibble(rownames = "envvar") %>% 
  mutate(across(starts_with("NMDS"),
                ~ .x * ordiArrowMul(pred_envfit)))
envfit_facs <- scores(pred_envfit, 
                      display = "factors",
                      choices = 1:3) %>% 
  as_tibble(rownames = "envvar") %>% 
  mutate(across(starts_with("NMDS"),
                ~ .x * ordiArrowMul(pred_envfit)))

# Hydrology plot
hydro_plot <- ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.8) +
  geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               data = envfit_vecs,
               linewidth = 1,
               color = "grey50",
               alpha = 0.6) +
  geom_text(aes(x = NMDS1, y = NMDS2, label = envvar),
            data = envfit_vecs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

# Ice plot
ice_facs <- filter(envfit_facs, str_detect(envvar, "ice_type"))
ice_plot <- ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.6) +
  geom_point(aes(x = NMDS1, y = NMDS2),
             data = ice_facs,
             size = 2) +
  ggrepel::geom_text_repel(aes(x = NMDS1, y = NMDS2, label = envvar),
                           data = ice_facs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

# Macrozooplankton plot
zoop_facs <- filter(envfit_facs, str_detect(envvar, "zoop"))
zoop_plot <- ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust), alpha = 0.6) +
  geom_point(aes(x = NMDS1, y = NMDS2),
             data = zoop_facs,
             size = 2) +
  ggrepel::geom_text_repel(aes(x = NMDS1, y = NMDS2, label = envvar),
                           data = zoop_facs) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic()

layout <- "
AAAA
AAAA
BBCC
"
hydro_plot + ice_plot + zoop_plot + 
  plot_layout(design = layout, guides = "collect")
```
